---
title: "R Notebook"
output: html_notebook
---

```{r libs}
library(tidyverse)
library(rjags)
library(coda)
```

## Simple binary latent class model.

Let $\alpha_i \in {1,0}$ (mastery, non-mastery)

Let $Y_{ij} \in {1,0}$ (correct, incorrect)

Let $g_j \in [0,1]$ is the probability of a correct response $p(Y_{ij}=1|\alpha_i=0)$ (guessing parameter)

Let $s_j = p(Y_{ij}=0|\alpha_i=1)$ (slipping parameter).

$${\cal L}({\bf Y}| {\bf \alpha}, {\bf g}, {\bf s}) =
  \prod_{i=1}^N \prod_{j=1}^J p(Y_{ij}|\alpha_i,g_j,s_j) = 
  \prod_{i=1}^N \prod_{j=1}^J \left [
  \begin{array}[l]
  \ Y_{ij}\left ((1-s_j)\alpha_i + (1-\alpha_i)*g_j\right ) \ +\\
  (1-Y{ij})\left (1-(1-s_j)\alpha_i - (1-\alpha_i)*g_j\right )
  \end{array}
  \right ]
$$

## Jags Model, first try

```{r model1}
write("
# data is Y[N,J];
var alpha[N], pij[N,J];  
model {
  for (i in 1:N) {
    alpha[i] ~ dbern(palpha)
    pij[i,] <- (1-s[])*alpha[i]+g[]*(1-alpha[i])
    for (j in 1:J) {
      Y[i,j] ~ dbern(pij[i,j])
    }
  }
  for (j in 1:J) {
    s[j] ~ dbeta(.5,.5)
    g[j] ~ dbeta(.5,.5)
  }
  palpha ~ dbeta(.5,.5)
}", "lc2model1.jags")
```

## Load ACED data

```{r loadACED}
ACED <- read_csv("https://ecd.ralmond.net/ACED/ACED-outcomes.csv")
CRitems <- ACED %>% select(matches("tCommonRatio")) %>% 
  mutate(across(starts_with("t"),~ as.numeric(.x =="Yes")))
head(CRitems)
```

```{r JAGSData}
CR.data <- list(Y=as.matrix(CRitems), J=ncol(CRitems), N=nrow(CRitems))
```

## JAGS Run

```{r SetupModel}
load.module("dic")
lc2.model1 <- jags.model("lc2model1.jags",CR.data,
                             n.chains=5)
```

```{r sampleModel1}
model1.samp <- coda.samples(lc2.model1,c("palpha","g","s","deviance"), 2000)
                            
```

```{r plotIt}
plot(model1.samp[,c("palpha","s[1]","g[1]")])
```

```{r summary}
summary(model1.samp)
```

## Again with some initial values.

```{r modelInits}
inits=list(list(palpha=.2),list(palpha=.2),list(palpha=.5),list(palpha=.8),list(palpha=.8))
lc2.model1i <- jags.model("lc2model1.jags",CR.data,inits,n.chains=5)
```

```{r sampleModel1i}
model1i.samp <- coda.samples(lc2.model1i,c("palpha","g","s","deviance"), 2000)
                            
```

```{r plotIt1i}
plot(model1i.samp[,c("palpha","s[1]","g[1]")])
```

## Different data set

```{r explicit}
ExplicitItems <-  ACED %>% select(matches("tExplicitGeometric")) %>% 
  mutate(across(starts_with("t"),~ as.numeric(.x =="Yes")))
summary(CRitems)
summary(ExplicitItems)
```

```{r ExpData}
Exp.data <- list(Y=as.matrix(ExplicitItems), J=ncol(ExplicitItems), N=nrow(ExplicitItems))
```

## JAGS Run

```{r SetupModele}
lc2.model1e <- jags.model("lc2model1.jags",Exp.data,
                             n.chains=5)
```

```{r sampleModel1e}
model1e.samp <- coda.samples(lc2.model1e,c("palpha","g","s","deviance"), 2000)
                            
```

```{r plotIt1e}
plot(model1e.samp[,c("palpha","s[1]","g[1]")])
```

## With inits to setup multiple modes.

```{r SetupModelei}
lc2.model1ei <- jags.model("lc2model1.jags",Exp.data,inits,
                             n.chains=5)
```

```{r sampleModel1ei}
model1ei.samp <- coda.samples(lc2.model1ei,c("palpha","g","s","deviance"), 2000)
```
```{r plotIt1ei}
plot(model1ei.samp[,c("palpha","s[1]","g[1]")])
```


## Alternate models:

Want probability of success for masters to always be higher than for non-masters.

$ (1-s_j) \geq g_j$

$ g_j' = \phi_j (1-s_j) $
$\phi_j \sim Beta(.5,.5)$

```{r model2}
write("
# data is Y[N,J];
var alpha[N], pij[N,J], g[J];  
model {
  for (i in 1:N) {
    alpha[i] ~ dbern(palpha)
    pij[i,] <- (1-s[])*(alpha[i]+phi[]*(1-alpha[i]))
    for (j in 1:J) {
      Y[i,j] ~ dbern(pij[i,j])
    }
  }
  for (j in 1:J) {
    s[j] ~ dbeta(.5,.5)
    phi[j] ~ dbeta(.5,.5)
    g[j] <- (1-s[j])*phi[j]
  }
  palpha ~ dbeta(.5,.5)
}", "lc2model2.jags")
```

## Post hoc reversal.

```{r headMCMC}
arow <- model1ei.samp[[1]][1,]
gvars <- paste("g[",1:6,"]",sep="")
svars <- paste("s[",1:6,"]",sep="")
arow
xrow <- arow
xrow["palpha"] <- 1-arow["palpha"]
xrow[gvars] <- 1-arow[svars]
xrow[svars] <- 1-arow[gvars]
xrow
```
