---
title: "Beta and Dirichlet Parameters"
author: "Russell Almond"
date: "January 24, 2019"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
library(shiny)
knitr::opts_chunk$set(echo = TRUE)
```

## Beta Distribution

Consider two [gamma](GammaParams.Rmd) random variables, $A$ with shape parameter $\alpha$ and $B$ with shape parameter $\beta$ with the same scale or rate parameter. Now consider the ratio $\theta = \frac{A}{A+B}$. This variable has a [beta distribution](https://en.wikipedia.org/wiki/Beta_distribution). Note that this distribution has a support of $[0,1]$, so it is excellent choice for modelling probabilities.

$$ f(\theta) = \frac{\Gamma(\alpha + \beta)}{\Gamma(\alpha)\Gamma(\beta)} \theta^{\alpha -1}(1-\theta)^{\beta-1}$$ 

* Mean: $ E[\theta] = \frac{\alpha}{\alpha+\beta} $

-   Mode: $ \frac{\alpha-1}{\alpha+\beta-2}$

-   Var: $\text{Var}(\theta) = \frac{\alpha\beta}{(\alpha + \beta)^2(\alpha+\beta+1)}$

Let $n = \alpha+\beta$ and $p = \alpha/n$. This is an alternative parameterization in terms of mean, and *effective sample size*.

```{r beta, echo=FALSE}
inputPanel(
  sliderInput("alpha", label = "Shape parameter 1 (alpha)",
              min=0, max=15, value=3, step=.5),
  
  sliderInput("beta", label = "Shape parameter 2 (beta)",
              min = 0, max = 15, value = 3, step = .5)
)

renderPlot({
  alpha <- as.numeric(input$alpha)
  beta <- as.numeric(input$beta)
  curve(dbeta(x,alpha,beta),
        xlim=c(0,1),
        main=paste("Beta distribution with alpha =",alpha,
                   "and beta = ",beta),
        xlab="X",ylab="Density")

})
```

Note that for $\alpha=\beta=1$, the beta distribution is the uniform distribution over 0, 1.

There are three proposed beta distributions for "non-informative priors."

-   Uniform distribution: $\alpha=\beta=1$
-   Jeffrey's Rule Prior: $\alpha=\beta=1/2$. This is uniform after a logistic transformation.
-   Jaynes' Prior: $\alpha=\beta=0$. This is not a proper probability distribution, but is the limit of probability densities.

## Dirichlet Distribution

Consider $K$ gamma varaibles $A_1, \ldots, A_K$ with shape parameters $\alpha_1, \ldots, \alpha_K$, and the same scale parameter.

Now consider the dependent collection of random variables $\theta_1, \ldots, \theta_K$, where $\theta_k = A_k/(\sum_k A_k)$.

This random variable is defined over a $k$ dimensional simplex, as $\sum_k \theta_k = 1$.

This is call the Dirichlet distribution and it has the density function:

$$ f(\theta_1,\ldots,\theta_K) = \frac{\Gamma(\alpha_1 + \cdots + \alpha_K)}{\Gamma(\alpha_1)\cdots\Gamma(\alpha_K)} \theta_1^{\alpha_1 -1}\cdots\theta_K^{\alpha_K-1}$$ Note that each component follows a beta distribution with parmeters $\alpha=\alpha_k$ and $\beta=\sum_{k'\not=k} \alpha_{k'}$.

Once again, the sum of the alphas is an *effective sample size* and this distribution can be parameterized via its mean vector and effective sample size.

```{r ternay}
# github:bursettles/ternary.R
library(klaR)

cart2ternary <- function(x, y) {
  b = y + 1/3
  a = -0.5*(x * 3 / sqrt(3) - 1 + b)
  c = 1 - (a + b)
  cbind("a" = a, "b" = b, "c" = c)
}

triimage <- function(x, y = NULL, z = NULL, main = "", label = 1:3, col = heat.colors(100), grid = seq(0.1, 0.9, by = 0.1),
                     useRaster = TRUE, set.par = TRUE, ...) {
  # setup the plot
  # this is necessary to avoid problems cause by using triplot to setup the graph;
  # see description of set.par in triplot()
  if (set.par) {
    if (main != "") 
        newmar <- c(0, 0, 4, 0) + 0.1
    else newmar <- rep(0, 4) + 0.1
    opar <- par(mar = newmar)
    on.exit(par(opar))
  }

  triplot(main = main, label = FALSE, frame = FALSE, grid = FALSE, set.par = FALSE, ...)
  
  # draw the image
  if (missing(z)) {
    image(x, col=col, add = TRUE, useRaster = useRaster)
  } else {
    image(x, y, z, col=col, add = TRUE, useRaster = useRaster)
  }
  
  # Draw the frame and grid on top of the image
  triframe(label = label)
  invisible(trigrid(grid))
}

outer.ternary <- function(fun, length=251, ...) {
  factory <- function(f, ...) {
    function(x, y) {
      triple = cart2ternary(x, y)
      nas = apply(triple, 1, function(x) { !all(x >= (0 - 1e-7) & x <= (1 + 1e-7)) } )
      triple[nas,] = c(NA, NA, NA)
      f(triple[,1], triple[,2], triple[,3], ...)
    }
  }
  
  x = seq(-sqrt(3)/3, sqrt(3)/3, length=length)
  y = seq(-1/3, 2/3, length=length)
  z = outer(x, y, factory(fun, ...))
  list(x=x, y=y, z=z)
}

entropy <- function(a, b, c) {
  -a * log(a) - b * log(b) - c * log(c)
}

hardness <- function(a, b, c) {
  1 - (a*a + b*b + c*c)
}
  
lc <- function(a, b, c) {
  1 - apply(cbind(a,b,c), 1, max)
}

margin <- function(a, b, c) {
  m <- function(row) {
    s = sort(row, decreasing = TRUE)
    s[2] - s[1]
  }
  apply(cbind(a, b, c), 1, m)
}

nlmp <- function(a, b, c) {
  -log(apply(cbind(a,b,c), 1, max))
}

##par(mfrow=c(2,3))
col = rainbow(100)
grid = seq(0.2, 0.8, by = 0.2)
# triimage(outer.ternary(entropy, length=101), useRaster = TRUE, main = "Entropy", col = col, grid = grid)
# triimage(outer.ternary(hardness, length=101), useRaster = TRUE, main = "Hardness", col = col, grid = grid)
# triimage(outer.ternary(margin, length=101), useRaster = TRUE, main = "Margin", col = col, grid = grid)
# triimage(outer.ternary(lc, length=101), useRaster = TRUE, main = "LC", col = col, grid = grid)
#triimage(outer.ternary(nlmp, length=101), useRaster = TRUE, main = "NLMP", col = col, grid = grid)

alphas <- c(3,2,1)
ddirich <- function (a,b,c) MCMCpack::ddirichlet(cbind(a,b,c),alphas)
triimage(outer.ternary(ddirich,length=101), useRaster = TRUE,
         main=paste("Dirichlet distribution with parameters: ",
                   paste(alphas,collapse=", ")), 
         col=col,grid=grid)
```
