---
title: "Gellman and Hill 13.2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nlme)

```

_Models for adjusting individual ratings:  a committee of 10 persons is evaluating 100 job applicants.  Each person on the committee reads 30 applications (structured so that each application is read by 3 people) and gives a numerical rating between 1 and 10._

_(a) It would be nautal to rate the applications based on their combined scores; however, there is aworry that different raters use different standards (severity) and we would like to correct for this.  Set up a model for the ratings (with parameters for the applicants and the raters)._

### Problem setup 
I <- 100L # applicants
J <- 10L  # raters
W <- 3L   # ratings per applicant

theta[i] -- Ability of Applicant i
theta ~ unif(1,10)  _1 and 10 are hyperparameters_

eta[j] -- Severity of Rater j
eta ~ norm(0,tau) _0 and tau are hypeparameters_
tau <- 2

epsilon[i,j] -- measurement error
epsilon ~ norm(0,sigma)
sigma <- 1

Y[i,j] = theta[i]+eta[j] +epsilon[i,j]

Score ~ (1 | applicant) + (1 | rater)

### Design Matrix

There are `r choose(10,3)` ways of assigning 3 raters, so more patterns than we have rows.

```{r}
I <- 100L # applicants
J <- 10L  # raters
W <- 3L   # ratings per applicant
tau <- 2  # SD of raters
sigma <- 1 # Error SD

assignment <- matrix(0L,I,J)
for (i in 1L:I) {
  workload <- colSums(assignment)
  available <- which (workload < W*I/J)
  if (i > 75L)
    cat("Round ",i,": available = ",
        paste(available,collapse=", "),"\n")
  ## If not W raters available, then swap until we have enough free raters.
  while (length(available) < W) {
    slacker <- which.min(workload)
    pswaps <- which(!assignment[1L:(i-1L),slacker])
    swaprow <- sample(pswaps,1L)
    swapcol <- sample(which(as.logical(assignment[swaprow,])),1L)
    assignment[swaprow,swapcol] <- 0L
    assignment[swaprow,slacker] <- 1L
    workload <- colSums(assignment)
    available <- which(workload < W*I/J)
    cat("Round ",i,"x: availble=",paste(available,collapse=", "),
        "\n")
  } 
  assignment[i,sample(available,W)] <- 1L
}
colSums(assignment)
rowSums(assignment)
write.csv(assignment,"assignment.csv")
```

## Simulating parameters
ability[i] -- random value between 1 and 10
severity[j] -- random value with mean 0 and SD tau=2
rating[i,j] = ability[i] + severity[j] + error[i,j]



```{r}
## simulate parameters/latent variables
ability <- runif(I,1,10)
severity <- rnorm(J,0,tau)
applicant <- rep(1L:I,each=W)
rater <- 
  sapply(1L:I,
              function (i) 
                which(as.logical(assignment[i,])))
str(rater)
rating <- ability[applicant] + severity[rater] + rnorm(I*W,0,sigma)
rating <- pmax(1,pmin(rating,10)) ## between 1 and 10
ratings.df <- data.frame(applicant=applicant,
                         rater=as.vector(rater), 
                         rating=rating)
ratings.df
write.csv(ratings.df,"ratings.csv")

```
# Plot the data to see if it worked
```{r}
library(lattice)
ratings.df1 <- data.frame(ratings.df, ability=ability[applicant], severity=severity[rater])

```

```{r}
xyplot(rating~ability,data=ratings.df1)
xyplot(rating~ability|rater,data=ratings.df1)
boxplot(rating~rater,data=ratings.df1)
```



Fit the model
```{r}
library(arm)
fit <- lmer(rating ~ (1|applicant) + (1|rater), data=ratings.df)

display(fit)
sqrt(9^2/12)  # sd of unif(1,10)

```

## Check our simulation.

```{r}
plot(ability,coef(fit)$applicant[,1])
plot(severity,ranef(fit)$rater[,1])
plot(fit)
boxplot(resid(fit)~as.vector(rater))
```



_(b) It is possible that some persons on the committee show more variation than others in their ratings.  Expand your model to allow this._

Change this part of the model:
epsilon[i,j] -- measurement error
epsilon ~ norm(0,sigma[j])
sigma[j] ~ gamma(alpha,scale)
alpha <- 2
scale <- .5
```{r}
alpha <- 2
scale <- .5
curve(dgamma(x,alpha,scale=scale),xlim=c(0,5))
```

Simulation is almost the same, and we can reuse the assignment matrix.

```{r}
## simulate parameters/latent variables
#ability <- runif(I,1,10)
#severity <- rnorm(J,0,tau)
#applicant <- rep(1L:I,each=W)
#rater <- 
#  sapply(1L:I,
#             function (i) 
#                which(as.logical(assignment[i,])))
#str(rater)
sigma2 <- rgamma(J,alpha,scale=scale)
rating2 <- ability[applicant] + severity[rater] + rnorm(I*W,0,sigma2[rater])
rating2 <- pmax(1,pmin(rating2,10)) ## between 1 and 10
ratings2.df <- data.frame(applicant=applicant,
                         rater=as.vector(rater), 
                         rating=rating2,
                         severity=severity[rater],
                         ability=ability[applicant],
                         sigma2=sigma2[rater])
ratings2.df
```

```{r}
xyplot(rating~ability,data=ratings2.df)
xyplot(rating~ability|rater,data=ratings2.df)
boxplot(rating~rater,data=ratings2.df)

```

Go ahead and fit the wrong model (we know this doesn't have the right variance structure).
```{r}
fit2 <- lmer(rating ~ (1|applicant) + (1|rater), data=ratings2.df)

display(fit2)
```
```{r}
plot(ability,coef(fit2)$applicant[,1])
plot(severity,ranef(fit2)$rater[,1])
plot(fit2)
boxplot(resid(fit2)~as.vector(rater))

```

