---
title: "R Notebook"
output: html_notebook
---

```{r libraries}
library(rjags)
library(coda)
```

## JAGS from the Command Line

The command line version of JAGS uses the following files:

1.  *Command File* --- Set of instructions on what to do:
2.  *Model (Bug) file* -- description of the model.
3.  *Data* (as an R list)
4.  *Initial values* (as an R list)

## Example LSAT

Responses from 1000 examinees on 5 items from the LSAT.

Will use a 1PL model:

$$\Pr(X_{jk}=1|\theta_i, \alpha_i, \beta) = 
\text{logit}^{-1} \beta \theta_j - \alpha_k$$ Need priors for

Discrimination $\beta \sim |N(0,10)|$

Difficulties $\alpha_k \sim N(0,10)$

Ability (Population Distribution) $\theta_j \sim N(0,1)$

### LSAT Command File

```{jags}
model in lsat.bug
data in lsat-data.R
load glm
compile, nchains(2)
inits in lsat-init.R
initialize
samplers to foo-samplers.txt
update 1000 
monitor alpha 
monitor beta 
update 2000 
coda *
```

[test1.cmd](lsat/test1.cmd)

### LSAT Model File

```{r lsat.jags}
lsat.jags <-
"var r[N,K];
data {
   for (j in 1:culm[1]) {
      r[j,] <- response[1, ];
   }
   for (i in 2:R) {
      for (j in (culm[i-1] + 1):culm[i]) {
         r[j,] <- response[i, ];
      }
   }
}
model {
# Rasch model
   for (j in 1:N) {
      for (k in 1:K) {
         logit(p[j,k]) <- beta*theta[j] - alpha[k];
         r[j,k] ~ dbern(p[j,k]);
      }
      theta[j] ~ dnorm(0,1);
   }
# Priors
   for (k in 1:K) {
      alpha[k] ~ dnorm(0,0.0001);
      a[k] <- alpha[k] - mean(alpha[]);
   }
   beta ~ dnorm(0,0.0001) T(0,);

# Compute probability of response pattern i, for later use in computing G^2
  theta.new ~ dnorm(0,1);          # ability parameter for random student 
  for(k in 1:K) {
       logit(p.theta[k]) <- beta*theta.new - alpha[k];  
       for(i in 1:R) {
          p.item[i,k] <- p.theta[k]^response[i,k] * (1-p.theta[k])^(1-response[i,k]);
       }
  } 
  for(i in 1:R) {    
     # P_i|theta = PROD_k p_k|theta   
     P.theta[i] <- prod(p.item[i,]);
  }
}"

```

[lsat.bug](lsat/last.bug)

#### Data Block

```{declaration}
var r[N,K];
```
Declares a new local variable a N=100 by K=5 matrix.

The code in the data block is run only once.
In this case, it unpacks the `response` (32x5) into `r` (1000x5)
```{dataBlock}
data {
   for (j in 1:culm[1]) {
      r[j,] <- response[1, ];
   }
   for (i in 2:R) {
      for (j in (culm[i-1] + 1):culm[i]) {
         r[j,] <- response[i, ];
      }
   }
}
```


#### Model block

The model block almost always has a section for the likelihood and for the prior.  Sometimes it also contains the calculation of some additional values used for reporting.

```{model}
model {
  ...
}
```

##### Likelihood
Note that in JAGS models need to loop over array indexes.

```{RaschModel}
## Rasch model
  for (j in 1:N) { ## Loop over examinees
    for (k in 1:K) { ## Loop over items
      logit(p[j,k])) <- beta*theta[j] - alpha[k];
      r[j,k] ~ dbern(p[j,k]);
    }
    theta[j] ~ dnorm(0,1);
  }
```

##### Priors

Note that JAGS uses `~ dXXX` for priors.

`dnorm` takes mean and _precision_ (1/variance) as parameters.  So for SD=10 -> $\sigma^{-2}=.0001$

`T(0,)` truncates the normal distribution, so this gives us a 1/2 normal prior.

```{Priors}
  for (k in 1:K) {
    alpha[k] ~ dnorm(0,0.0001);
    a[k] <- alpha[k] - mean(alpha[]);
  }
  beta ~dnorm(0,0.0001)T(0,);
```

`a` is a centered version of `alpha`.  We can compute it here, or as a post processing step.


### Data Values

These are the output of `save`, so they look a bit odd.

[lsat/lsat-data.R](lsat/lsat-data.R)

```{r inputdata}
lsat.dat <- read.jagsdata("lsat/lsat-data.R")
data.frame(lsat.dat$response, lsat.dat$m, lsat.dat$culm)
lsat.dat$K <- lsat.dat$T
lsat.dat$T <- NULL
```

Note: Data are in a compressed format. So first step is to replicate out to a 1000 x 5 matrix.

### Initial Values

```{r InitialValues}
lsat.inits <- rjags::read.jagsdata("lsat/lsat-init.R")
lsat.inits
```
Note that you can create separate inits for each chain if you want.

Values for which you don't have inits (e.g., theta) will have randomly drawn inits.

## Running using R

```{r setupJAGS}
#library(JAGS)
load.module("glm")
#last.dat <- read.jagsdata("lsat-data.R")
lsat.dat$m <- NULL
#lsat.inits <- read.jagsdata("lsat-init.R")
lsat.model <- jags.model("lsat/lsatA.bug", lsat.dat, lsat.inits, n.chains=2, n.adapt=500)
```

```{r runJAGS}
update(lsat.model, 500) ## Burn-in
lsat.samp <- coda.samples(last.model, c("alpha","beta"), n.iter=2000)
```


## Quick and dirty Coda

### Summary

```{r codaSummary}
summary(lsat.samp)
```

### Traceplots

```{r traceplots}
plot(lsat.samp)
```

### Autocorrelation

```{r acf1}
acfplot(lsat.samp)
```

```{r autocorr}
autocorr.diag(lsat.samp)
autocorr.plot(lsat.samp)
```

```{r autocorr}
autocorr(lsat.samp)
```

### Gelman--Rubin R

```{r GelmanRubinR}
gelman.diag(lsat.samp)
gelman.plot(lsat.samp)
```

