---
title: "Negative Binomial Parameters"
author: "Russell Almond"
date: "January 17, 2023"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(shiny)
library(ggplot2)
```

The [negative binomial distribution](https://en.wikipedia.org/wiki/Negative_binomial_distribution) can be thought of as a number of draws of non-black balls, $x$, from an urn with a proportion $p$ of black balls, until $r$ black balls have been drawn. When $r=1$, it is known as the *geometric distribution*

The probability of drawing exactly $x$ balls from an this urn is: $$ p(X=k|r,p) = \binom{k+r-1}{k} p^r (1-p)^{k}$$

The expected value is $\frac{r(1-p)}{p}$, and the variance is $\frac{r(1-p)}{p^2}$.

Sometimes this is written in terms of number of trials instead of number of failures (see Wikipedia entry).

```{r density, echo=FALSE}
inputPanel(
  sliderInput("r", label = "Number of successes:",
              min=0, max=50, value=5, step=1),
  
  sliderInput("p", label = "Probability of success:",
              min = 0, max = 1, value = .5, step = 0.01)
)

renderPlot({
  r <- as.numeric(input$r)
  p <- as.numeric(input$p)
  n <- qnbinom(.995,r,p)
  dat <- data.frame(x=0:n,y=dnbinom(0:n,r,p))
  ggplot(dat,aes(x,y)) +geom_col()  

})
```

Note that this distribution is positively skewed.

It also can come about from a Poisson process where the rate, $\lambda$ is a (gamma distributed) random variable. Used as an overdispersed Poisson.

Note the sum of negative binomials is also negative binomials.

Specifically, for geometric distribution: $\Pr( X>n | X>k) = \Pr(X > (n-k))$.
